---
title: 'Statistical Inference Project - Part 2'
author: 'Bhaskar S'
date: '14th May 2016'
output:
  pdf_document: default
  html_document:
    keep_md: yes
---
<br/>

## __Analysis of the ToothGrowth dataset in R__
<br/>

### __Synopsis__
<br/>

In this project, we analyze the __ToothGrowth__ dataset in R to perform basic Exploratory Data Analysis and Hypothesis Tests to compare the tooth growth by supplements and dosage.

<br/>

### __Exploratory Data Analysis__
<br/>

The __ToothGrowth__ dataset in R contains the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).

For this analysis, we will be using the R packages `data.table` and `ggplot2`. The following code segment loads the desired libraries:

```{r setup_1, echo = TRUE}
library(data.table)
library(ggplot2)
```

The following code segment loads the __ToothGrowth__ dataset that is built into R. This automatically creates a data frame called `ToothGrowth`:

```{r load_1, echo = TRUE}
data('ToothGrowth')
```

We initialize a `data.table` variable called __TG__ that will be referenced in this analysis. We use a `data.table` instead of the built-in data frame `ToothGrowth` for ease of use. Also, the `dose` data column is converted to a factor variable. Finally, we remove the data frame `ToothGrowth` once we initialize our `data.table` called __TG__ for efficiency. The following code segment performs the desired initializations:

```{r setup_2, echo = TRUE}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
TG <- data.table(ToothGrowth)
rm(ToothGrowth)
```

As a first step in the exploratory data analysis, we want to compactly display the structure and contents of our `data.table` called __TG__. The following code displays the structure and contents of the `data.table` called __TG__ using the R `str` function:

```{r display_1, echo = TRUE}
str(TG)
```

As we can see, the `data.table` called __TG__ has __60__ rows and __3__ columns (`len`, `supp`, `dose`). The columns `supp` and `dose` are factor variables and represent the _supplement_ and _dosage_ respectively.

The next step in the exploratory data analysis is to display the data in the `data.table` called __TG__ for visual analysis. We display the data in three sub-plots by `dose` with the `supp` along the x-axis and the `len` along the y-axis.

The following code plots the `data.table` called __TG__ as a _boxplot_:

```{r plot_4, echo = TRUE, fig.path = './figures/'}
ggplot(TG, aes(x = supp, y = len, fill = supp)) +
    geom_boxplot(aes(fill = supp)) +
    labs(x = 'Supplement', y = 'Length') +
    ggtitle('Supplement vs Length by Dosage') +
    facet_grid(~dose) +
    theme_bw() +
    theme(plot.title = element_text(face = 'bold', size = 10),
          axis.title = element_text(face = 'bold', size = 10))
```

From the above plot, we can infer the following facts:

* For `dose` __0.5__ and __1__, the _supplement_ __OJ__ seems to have a better growth result than the _supplement_ __VC__
* For `dose` __2__, however, the _supplement_ __VC__ seems to have a better growth result than the _supplement_ __OJ__

The next step in the exploratory data analysis is to display the `summary` statistics for the `len` data column in the `data.table` called __TG__.

For this, we initialize __6__ subsets of the `data.table` called __TG__ using the data columns `supp` and `dose`. The following code segment performs the desired initializations:

```{r setup_3, echo = TRUE}
TG1.OJ <- TG[supp == 'OJ' & dose == 0.5]
TG1.VC <- TG[supp == 'VC' & dose == 0.5]
TG2.OJ <- TG[supp == 'OJ' & dose == 1]
TG2.VC <- TG[supp == 'VC' & dose == 1]
TG3.OJ <- TG[supp == 'OJ' & dose == 2]
TG3.VC <- TG[supp == 'VC' & dose == 2]
```

The following code displays the the `summary` statistics for the `len` data column from each of the __6__ subsets we created above:

```{r display_2, echo = TRUE}
summary(TG1.OJ$len)
summary(TG1.VC$len)
summary(TG2.OJ$len)
summary(TG2.VC$len)
summary(TG3.OJ$len)
summary(TG3.VC$len)
```

From the above `summary` statistics, we can infer the following facts:

* For `dose` __0.5__ and __1__, the _supplement_ __OJ__ seems to have a better growth result than the _supplement_ __VC__
* For `dose` __2__, however, there does not seem to be any major advantage between _supplement_ __OJ__ or _supplement_ __VC__
<br/>

### __Hypothesis Testing__
<br/>

#### __Hypothesis Testing on Supplement Types__
<br/>

Looking at the data, we have two supplement types (OJ and VC) and the sample size is equal to __30__. Also, we do not have any knowledge of the population variance. As a result, we will be conducting a __t__ hypothesis test to find the __p-value__ using the R `t.test` function.

We will test the _null_ hypothesis that the mean tooth growth is equal between the supplements __OJ__ and __VC__ with a 95% __Confidence Interval__. Statistically, we are testing for __H0: mu1 = mu2__. The _alternate_ hypothesis is __Ha: mu1 != mu2__. Since the _null_ hypothesis is testing for equality, this is a __two-tail__ test. If the __p-value__ is < 0.05, we reject the _null_ hypothesis.

For this, we initialize __2__ subsets of the `data.table` called __TG__ using the data column `supp`. The following code segment performs the desired initializations:

```{r setup_4, echo = TRUE}
TG.OJ <- TG[supp == 'OJ']
TG.VC <- TG[supp == 'VC']
```

The following code performs a two-tailed __t__ hypothesis test:

```{r compute_1, echo = TRUE}
t.test(len ~ supp, data = TG, var.equal = FALSE, paired = FALSE, conf.level = 0.95)
```

From the above, we see the __p-value__ is greater than __0.05__ and hence we fail to reject the _null_ hypothesis and conclude with 95% confidence level that there is no major difference between the mean growths from the two supplements __OJ__ and __VC__.
<br/>

#### __Hypothesis Testing on Dosage Levels__
<br/>

Looking at the data, we have three dosage levels (0.5, 1, 2) and the sample size is equal to __20__ for each level. Also, we do not have any knowledge of the population variance. As a result, we will be conducting a __t__ hypothesis test to find the __p-value__ using the R `t.test` function.

We will test the _null_ hypothesis for each of the dosage levels that the mean tooth growth is equal between the supplements __OJ__ and __VC__ with a 95% __Confidence Interval__. Statistically, we are testing for __H0: mu1 = mu2__. The _alternate_ hypothesis is __Ha: mu1 != mu2__. Since the _null_ hypothesis is testing for equality, this is a __two-tail__ test. If the __p-value__ is < 0.05, we reject the _null_ hypothesis.

For this, we initialize __3__ subsets of the `data.table` called __TG__ using the data column `dose`. The following code segment performs the desired initializations:

```{r setup_5, echo = TRUE}
TG1 <- TG[dose == as.factor(0.5)]
TG2 <- TG[dose == as.factor(1)]
TG3 <- TG[dose == as.factor(2)]
```

The following code performs a two-tailed __t__ hypothesis test for dosage level __0.5__:

```{r compute_2, echo = TRUE}
t.test(len ~ supp, data = TG1, var.equal = FALSE, paired = FALSE, conf.level = 0.95)
```

From the above, we see the __p-value__ of __0.006359__ is less than __0.05__ and hence we reject the _null_ hypothesis and conclude with 95% confidence level that there is difference between the mean growths from the two supplements __OJ__ and __VC__ for dosage __0.5__.

The following code performs a two-tailed __t__ hypothesis test for dosage level __1__:

```{r compute_3, echo = TRUE}
t.test(len ~ supp, data = TG2, var.equal = FALSE, paired = FALSE, conf.level = 0.95)
```

From the above, we see the __p-value__ of __0.001038__ is less than __0.05__ and hence we reject the _null_ hypothesis and conclude with 95% confidence level that there is difference between the mean growths from the two supplements __OJ__ and __VC__ for dosage __1__.

Finally, the following code performs a two-tailed __t__ hypothesis test for dosage level __2__:

```{r compute_4, echo = TRUE}
t.test(len ~ supp, data = TG3, var.equal = FALSE, paired = FALSE, conf.level = 0.95)
```

From the above, we see the __p-value__ of __0.9639__ is greater than __0.05__ and hence we fail to reject the _null_ hypothesis and conclude with 95% confidence level that there is no difference between the mean growths from the two supplements __OJ__ and __VC__ for dosage __2__.
<br/>

### __Summary__
<br/>

We conclude the following facts from our analysis:

* The supplements on their own do not seem to have an influence on the tooth growth
* The dosage levels seems to have an influence on the tooth growth
